\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-19 analysis/paper/paper.Rmd

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: >
  The abstract of the article. *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, data visualization]
  plain:     [survival analysis, absolute risk, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The grey area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments). The points shown in the figure are select person-moments from the case and base series. We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

@@

# dev.off()
layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xmax = max(new_time),
     # xlim = range(new_time), 
     conf.int = FALSE, col = myCols, 
     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
     #                exp(coef(cox_model)), 
     #                exp(confint(cox_model))[1], 
     #                exp(confint(cox_model))[2]), 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Casebase)",
                  "Screening (Cox)", "Screening (Casebase)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)

@@

As above, we can look at the population-time plot for a graphical summary of the event incidence. Here, we colour the exposed person-time (i.e. after transplant) in a darker shade of gray. As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.

```{r stanford-poptime, eval = eval_cs4}
stanford_popTime <- popTime(jasa,
  time = "futime",
  event = "fustat"
)
# plot(stanford_popTime,
#      ribbon.params = list(fill = "grey50")) + 
#   paper_gg_theme

@@

plot(stanford_popTime,
     ribbon.params = list(fill = "grey50")) + 
    geom_segment(data = tx_exp,
                 aes(x = x, xend = original.time,
                     y = ycoord, yend = ycoord),
                 colour = "grey30", size = 1) + 
    geom_point(data = filter(stanford_popTime, event == 1),
               aes(x = time, y = yc, color = "Case series",
                   fill = "Case series"),
               size = 1,
               alpha = 1,
               shape = 21) +
  paper_gg_theme
```

@@

```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit", message = FALSE, results="hide"}
plot(fit3, hazard.params = list(xvar = "futime",
                                by = "exposure",
                                alpha = 0.05,
                                ylab = "Hazard",
                                data = cb_data))
```

The non-proportionality seems to be more pronounced at the beginning of follow-up than the end. Finally, we can turn these estimates of the hazard function into estimates of the cumulative incidence functions.

```{r stanford-risk, eval = eval_cs4, echo = TRUE, dependson="stanford_fit"}
# Compute absolute risk curves
newdata <- data.frame(exposure = c(0, 1))
absrisk <- absoluteRisk(fit3,
  newdata = newdata,
  time = seq(0, 425, length.out = 100)
)

@@

<!-- Note that we can easily adapt the code above to the situation where a patient receives a heart transplant at a point in time of interest, for example after 1 month. -->

```{r stanford-1y-haz, eval = FALSE, echo = FALSE}
# Compute hazards---
# First, create a list of time points for both exposure status
one_yr_haz <- data.frame(futime = seq(0, 365,
  length.out = 100
))
one_yr_haz <- mutate(one_yr_haz,
  offset = 0,
  exposure = if_else(futime < 1, 0, 1)
)
one_yr_haz$hazard <- exp(predict(fit4,
  newdata = one_yr_haz,
  type = "link"
))
ggplot(one_yr_haz, aes(futime, hazard)) +
  geom_line() +
  paper_gg_theme +  
  ylab("Hazard") +
  xlab("Follow-up time (in days)") +
  geom_vline(xintercept = 1, linetype = "dashed")
```

<!-- We can then compare the 1-year mortality risk without transplant and with transplant at 1 month. -->

```{r stanford-1y-risk, eval = FALSE, echo = FALSE}
absoluteRisk(fit4,
  newdata = data.frame(exposure = 0),
  time = 365
)

# Use the trapezoidal rule to estimate the cumulative hazard
cumhaz_est <- pracma::trapz(one_yr_haz$futime, one_yr_haz$hazard)
1 - exp(-cumhaz_est)
```

<!-- As we can see, the risk estimate at 1-year is about 30% lower if the patient receives a heart transplant at 1 month. -->

<!-- We can also plot the hazard ratio: -->

```{r, eval = FALSE, dependson="stanford_fit"}
newtime <- seq(0, 425, length.out = 100)

# reference category
newdata <- data.frame(exposure = 0,
                      futime = newtime)

plot(fit3,
     type = "hr",
     newdata = newdata,
     var = "exposure",
     increment = 1,
     xvar = "futime",
     xlab = "Follow-up time (in days)",
     ci = FALSE,
     rug = TRUE)
abline(h = 1, lty = 2)
```

@@

```{r, echo=FALSE, eval = FALSE}
knitr::kable(data.frame(
  `feature` = c("model type", "time", "cumulative incidence", "non-proportional hazards", "model testing", "competing risks", "prediction"),
  `Cox` = c("semi-parametric", "left hand side of the equation", "step function", "interaction of covariates with time", " ", "difficult", "Kaplan-Meier-based"),
  `Case Base Sampling` = c(
    "fully parametric (logistic/multinomial regression)", "right hand side - allows flexible modeling of time",
    "smooth-in-time curve", "interaction of covariates with time", "make use of GLM framework (LRT, AIC, BIC)",
    "cause-specific cumulative incidence functions (CIFs) directly obtained via multinomial regression", "ROC, AUC, risk reclassification probabilities"
  )
),
col.names = c("Feature", "Cox model", "Case-base sampling"),
format = "latex", booktabs = TRUE,
caption = "Comparison between the Cox model and case-base sampling",
escape = FALSE
)
```

@@

<!-- # Appendix -->

<!-- A description of the variables can be found in Table \ref{tab:support1}, and a breakdown of each categorical variable appears in Table \ref{tab:support2}.  -->

<!-- \begin{table}[ht] -->
<!-- \centering -->
<!-- \begin{tabular}{ccccc} -->
<!-- \hline -->
<!-- \textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\ -->
<!-- \hline -->
<!-- age           & Age                                      &                 & double           & 0            \\ -->
<!-- death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\ -->
<!-- sex           &                                          & 2               & integer          & 0            \\ -->
<!-- slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\ -->
<!-- d.time        & Days of Follow-Up                        &                 & double           & 0            \\ -->
<!-- dzgroup       &                                          & 8               & integer          & 0            \\ -->
<!-- dzclass       &                                          & 4               & integer          & 0            \\ -->
<!-- num.co        & number of comorbidities                  &                 & double           & 0            \\ -->
<!-- edu           & Years of Education                       &                 & double           & 202          \\ -->
<!-- income        &                                          & 4               & integer          & 349          \\ -->
<!-- scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\ -->
<!-- avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\ -->
<!-- race          &                                          & 5               & integer          & 5            \\ -->
<!-- sps 	        & support physiology score day 3           &	               & double           &	1            \\ -->
<!-- aps           & APS III no coma, imp bun,uout for ph1,D3 &	               & double	          & 1            \\ -->
<!-- hday	        & Day in Hospital at Study Admit	         &                 & double           & 0            \\	 -->
<!-- diabetes      &	Diabetes (Com 27-28, Dx 73)	             &                 & double	          & 0            \\ -->
<!-- dementia      & Dementia (Comorbidity 6)                 &	               & double	          & 0            \\ -->
<!-- ca		        & Cancer State                             & 3               & integer	        & 0            \\ -->
<!-- meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\ -->
<!-- wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\ -->
<!-- hrt           & Heart Rate Day 3                         &                 & double           & 0            \\ -->
<!-- resp          & Respiration Rate Day 3                   &                 & double           & 0            \\ -->
<!-- temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\ -->
<!-- pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\ -->
<!-- alb           & Serum Albumin Day 3                      &                 & double           & 378          \\ -->
<!-- bili          & Bilirubin Day 3                          &                 & double           & 297          \\ -->
<!-- crea          & Serum creatinine Day 3                   &                 & double           & 3            \\ -->
<!-- sod           & Serum sodium Day 3                       &                 & double           & 0            \\ -->
<!-- ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\ -->
<!-- glucose       & Glucose Day 3                            &                 & double           & 470          \\ -->
<!-- bun           & BUN Day 3                                &                 & double           & 455          \\ -->
<!-- urine         & Urine Output Day 3                       &                 & double           & 517          \\ -->
<!-- adlp          & Activiites of Daily Living (ADL) Patient Day 3 &           & double           & 634          \\ -->
<!-- adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0            -->
<!-- \end{tabular} -->
<!-- \caption{Description of each variable in the SUPPORT dataset.} -->
<!-- \label{tab:support1} -->
<!-- \end{table} -->

<!-- \begin{table}[ht] -->
<!-- \centering -->
<!-- \begin{tabular}{cc} -->
<!-- \hline -->
<!-- \textbf{Variable} & \textbf{Levels}                      \\ -->
<!-- \hline -->
<!-- sex      & female                                        \\ -->
<!--          & male                                          \\ -->
<!-- dzgroup  & ARF/MOSF w/Sepsis                             \\ -->
<!--          & COPD                                          \\ -->
<!--          & CHF                                           \\ -->
<!--          & Cirrhosis                                     \\ -->
<!--          & Coma                                          \\ -->
<!--          & Colon Cancer                                  \\ -->
<!--          & Lung Cancer                                   \\ -->
<!--          & MOSF w/Malig                                  \\ -->
<!-- dzclass  & ARF/MOSF                                      \\ -->
<!--          & COPD/CHF/Cirrhosis                            \\ -->
<!--          & Coma                                          \\ -->
<!--          & Cancer                                        \\ -->
<!--          & \textgreater{}\$50k                           \\ -->
<!-- race     & white                                         \\ -->
<!--          & black                                         \\ -->
<!--          & asian                                         \\ -->
<!--          & other                                         \\ -->
<!--          & hispanic                                      \\ -->
<!-- ca       & Yes                                           \\ -->
<!--          & No                                            \\ -->
<!--          & Metastatic                                    \\ -->
<!-- \end{tabular} -->
<!-- \caption{Description of each level within each categorical variable in the SUPPORT dataset.} -->
<!-- \label{tab:support2} -->
<!-- \end{table} -->

<!-- In Figure \ref{fig:brier1} (A), we can see that all Cox models have similar average survival curves. On the other hand, Figure \ref{fig:brier1} (B) shows that the full model has the lowest Brier score overall. For this reason, we will use this model in our comparison between regularized Cox regression, Kaplan-Meier estimation, and the approach from \pkg{casebase}. -->

<!-- ```{r baselinePlotsPrep, eval=eval_cs3, echo = FALSE} -->
<!-- # 1. Cox with only sps -->
<!-- coxSPS <- survival::coxph(Surv(time = d.time, event = death) ~ sps,  -->
<!--                           data = train, x = TRUE) -->
<!-- abcoxSPS <- survival::survfit(coxSPS, newdata = test) -->
<!-- # 2. Cox with only aps -->
<!-- coxAPS <- survival::coxph(Surv(time = d.time, event = death) ~ aps,  -->
<!--                           data = train, x = TRUE) -->
<!-- abcoxAPS <- survival::survfit(coxAPS, newdata = test) -->
<!-- # 3. Cox with everything but sps and aps -->
<!-- coxFull <- survival::coxph(Surv(time = d.time, event = death) ~ . - aps - sps,  -->
<!--                            data = train, x = TRUE) -->
<!-- abcoxFull <- survival::survfit(coxFull, newdata = test) -->
<!-- # 4. Kaplan-Meier -->
<!-- KM <- survival::coxph(Surv(time = d.time, event = death) ~ 1, data = test,x=TRUE) -->
<!-- abKM <- survival::survfit(KM) -->
<!-- abKMcompare <- data.frame( -->
<!--   Time = abKM$time, -->
<!--   KM = 1 - abKM$surv -->
<!-- ) -->

<!-- # Combine all CI estimates from Cox models -->
<!-- baseCoxComparisonData <- data.frame( -->
<!--   Time = abcoxSPS$time, -->
<!--   SPS = 1 - rowMeans(abcoxSPS$surv), -->
<!--   APS = 1 - rowMeans(abcoxAPS$surv), -->
<!--   Full = 1 - rowMeans(abcoxFull$surv) -->
<!-- ) -->
<!-- ``` -->

<!-- ```{r baselines, echo=FALSE, eval=eval_cs3} -->
<!-- cox_ci <- ggplot(baseCoxComparisonData, aes(Time)) + -->
<!--   geom_step(aes(y = APS, colour = "APS")) + -->
<!--   geom_step(aes(y = Full, col = "Full")) + -->
<!--   geom_step(aes(y = SPS, colour = "SPS")) + -->
<!--   geom_step(data = abKMcompare, mapping = aes(x = Time, y = KM, color = "K-M")) + -->
<!--   labs(y = "Probability of Death", x = "Follow-up time (Years)", color = "Models") + -->
<!--   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) +  -->
<!--   paper_gg_theme +  -->
<!--   scale_x_continuous(labels = scaleFUNx, breaks = round(seq(0, 5, by = 1))) +  -->
<!--   scale_y_continuous(labels = scaleFUNy, n.breaks = 8) -->
<!-- ``` -->

<!-- ```{r brierpt1, echo=FALSE, eval=eval_cs3, fig.cap="\\label{fig:brier1} Comparison of three Cox regression models to the unadjusted Kaplan-Meier curve. (A) Cumulative Incidence curves. For a given model, the plotted curve is the average curve of each individual in the study. Each follow-up time seen in the study was used. There are minimal differences between the curves. APS coincides with the SPS curve, as they are considerably overlapped.(B) Time-dependent Brier scores, where a lower score is better. The full model performs best overall."} -->
<!-- # Extract time points -->
<!-- times <- sort(unique(test$d.time)) -->
<!-- # Compute Brier scores -->
<!-- brierAPS <- Score(list("PredictionModel" = coxAPS), data = test,  -->
<!--                   formula = Hist(d.time, death != 0) ~ 1, summary = "ipa",  -->
<!--                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
<!-- brierSPS <- Score(list("PredictionModel" = coxSPS), data = test,  -->
<!--                   formula = Hist(d.time, death != 0) ~ 1, summary = "ipa",  -->
<!--                   se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->

<!-- # First fix NA coefficients, then compute Brier score -->
<!-- coxFull$coefficients[which(is.na(coxFull$coefficients))] <- 0 -->
<!-- brierFull <- Score(list("PredictionModel" = coxFull), data = test,  -->
<!--                    formula = Surv(d.time, death != 0) ~ 1, summary = "ipa",  -->
<!--                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->
<!-- brierNull <- Score(list("PredictionModel" = KM), data = test,  -->
<!--                    formula = Surv(d.time, death != 0) ~ 1, summary = "ipa",  -->
<!--                    se.fit = FALSE, metrics = "brier", contrasts = FALSE, times = times) -->

<!-- # Combine all results for plotting -->
<!-- brierPlotter <- rbind( -->
<!--   data.frame(brierScore = brierAPS$Brier$score$Brier[brierAPS$Brier$score$model == "PredictionModel"],  -->
<!--              times = times, model = "APS"), -->
<!--   data.frame(brierScore = brierFull$Brier$score$Brier[brierFull$Brier$score$model == "PredictionModel"],  -->
<!--              times = times, model = "Full"), -->
<!--   data.frame(brierScore = brierSPS$Brier$score$Brier[brierSPS$Brier$score$model == "PredictionModel"],  -->
<!--              times = times, model = "SPS"), -->
<!--     data.frame(brierScore = brierNull$Brier$score$Brier[brierNull$Brier$score$model == "PredictionModel"],  -->
<!--              times = times, model = "K-M") -->
<!-- ) -->

<!-- cox_brier <- ggplot(data = brierPlotter, mapping = aes(x = times, y = brierScore, col = model)) + -->
<!--   geom_line() + -->
<!--   xlab("Follow-up time (Years)") + -->
<!--   ylab("Brier score") + -->
<!--   labs(color = "Models") + -->
<!--   # ggtitle("Brier scores over time") + -->
<!--   scale_color_manual(values = q4, breaks = c("APS", "Full", "SPS", "K-M")) +  -->
<!--   paper_gg_theme+ scale_x_continuous(labels=scaleFUNx,breaks = round(seq(0, 5, by = 1)))+scale_y_continuous(labels=scaleFUNy,n.breaks=8) -->

<!-- # Combine both plots using cowplot -->
<!-- plot_row <- cowplot::plot_grid(cox_ci + theme(legend.position = 'none'), -->
<!--                                cox_brier + theme(legend.position = 'none'), -->
<!--                                labels = c("A", "B")) -->
<!-- legend <- cowplot::get_legend( -->
<!--   # create some space to the left of the legend -->
<!--   cox_ci + theme(legend.box.margin = margin(0, 0, 0, 12)) -->
<!-- ) -->

<!-- cowplot::plot_grid(plot_row, legend, rel_heights = c(3, .4), nrow = 2) -->
<!-- ``` -->

>>>>

---
documentclass: jss
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba \AND'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title:
  formatted: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # If you use tex in the formatted title, also supply version without
  plain:     "casebase: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
  # For running headers, if needed
  short:     "\\pkg{casebase}: An Alternative Framework For Survival Analysis"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function. But to estimate risks, the baseline hazard needs to be estimated separately using a non-parametric method. This apparoch leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the casebase R package, which implements the ideas of Hanley & Miettinen through a simple and flexible interface. We describe how the case-base framework can also be used in more complex settings: competing risks, time-varying exposure, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through four different case studies.
  | *SRB and MT contributed equally to this work.
keywords:
  # at least one keyword must be supplied
  formatted: [survival analysis, absolute risk, "\\proglang{R}", data visualization]
  plain:     [survival analysis, absolute risk, R, data visualization]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}

@@

```{r plot-stratified-erspc-data, eval = eval_cs1, echo = FALSE, fig.width=7, fig.height=5, fig.asp=0.75, fig.cap="Population time plots for both treatment arms in the ERSPC dataset. The gray area can be thought of as N=88,232 (control group) and N=71,661 (screening group) rows of infinitely small rectangles (person-moments). The points shown in the figure are select person-moments from the case and base series. We see that more events are observed at later follow-up times, motivating the use of a non-constant in time hazard model."}
label_number_si <- function (accuracy = 1, unit = NULL, sep = NULL, ...) {
    sep <- if (is.null(unit)) 
        ""
    else " "
    scales:::force_all(accuracy, ...)
    function(x) {
        breaks <- c(0, 10^c(K = 3, M = 6, B = 9, T = 12))
        n_suffix <- cut(abs(x), breaks = c(unname(breaks), Inf), 
            labels = c(names(breaks)), right = FALSE)
        n_suffix[is.na(n_suffix)] <- ""
        suffix <- paste0(sep, n_suffix, unit)
        scale <- 1/breaks[n_suffix]
        scale[which(scale %in% c(Inf, NA))] <- 1
        number(x, accuracy = accuracy, scale = unname(scale), 
            suffix = suffix, ...)
    }
}
plot(pt_object, 
     add.base.series = TRUE,
     ratio = 0.7,
     facet.params = list(ncol = 2),
     ribbon.params = list(fill = "gray50"),
     case.params = list(size = 0.8),
     base.params = list(size = 0.8)) + 
  paper_gg_theme +  
  scale_y_continuous(labels = label_number_si()) + 
  ylab("Population Size")
```

@@

# dev.off()
layout(matrix(1:4, ncol = 2, byrow = FALSE), respect = FALSE)
par(mar = c(0, 4.1, 4.1, 0))
plot(survfit(cox_model, newdata = new_data),
     ylab = "",
     xlab = "",
     fun = "event",
     xmax = max(new_time),
     # xlim = range(new_time), 
     conf.int = FALSE, col = myCols, 
     # main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
     #                Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
     #                exp(coef(cox_model)), 
     #                exp(confint(cox_model))[1], 
     #                exp(confint(cox_model))[2]), 
     lwd = 2,
     yscale = 100,
     xaxt = 'n')
legend("topleft", 
       legend = c("Control (Cox)","Control (Case-base)",
                  "Screening (Cox)", "Screening (Case-base)"), 
       col = rep(myCols, each = 2),
       lty = c(1, 2, 1, 2), 
       lwd = 2,
       cex = 1.1,
       bg = "gray90")
lines(risks$exponential[,1], risks$exponential[,2], col = myCols[1], lty = 2, lwd = 2)
lines(risks$exponential[,1], risks$exponential[,3], col = myCols[2], lty = 2, lwd = 2)
text(x = max(new_time)*0.97, y = 0.05/100, labels = "(1)", cex = 1.5)

@@

As above, we can look at the population-time plot for a graphical summary of the event incidence (see Figure \ref{fig:cs4PopTime}. Here, we colour the exposed person-time (i.e. after transplant) in a darker shade of gray. As we can see, most events occur early during the follow-up period, and therefore we do not expect the hazard to be constant.

```{r stanford-poptime, eval = eval_cs4, fig.cap="\\label{fig:cs4PopTime} Population-time for the Stanford Transplant study. The darker shade of gray corresponds to the exposed person-time."}
stanford_popTime <- popTime(jasa,
  time = "futime",
  event = "fustat"
)
# plot(stanford_popTime,
#      ribbon.params = list(fill = "gray50")) + 
#   paper_gg_theme

@@

plot(stanford_popTime,
     ribbon.params = list(fill = "gray50")) + 
    geom_segment(data = tx_exp,
                 aes(x = x, xend = original.time,
                     y = ycoord, yend = ycoord),
                 colour = "gray30", size = 1) + 
    geom_point(data = filter(stanford_popTime, event == 1),
               aes(x = time, y = yc, color = "Case series",
                   fill = "Case series"),
               size = 1,
               alpha = 1,
               shape = 21) +
  paper_gg_theme
```

@@

```{r stanford-hazard, eval = eval_cs4, echo=TRUE, dependson="stanford_fit", message = FALSE, results="hide", fig.cap="\\label{fig:cs4hazardPlot} Plot of the hazard function for exposed and unexposed individuals."}
plot(fit3, hazard.params = list(xvar = "futime",
                                by = "exposure",
                                alpha = 0.05,
                                ylab = "Hazard",
                                data = cb_data))
```

Looking at Figure \ref{fig:cs4hazardPlot}, the non-proportionality seems to be more pronounced at the beginning of follow-up than the end. Finally, in Figure \ref{fig:cs4CIF}, we turn these estimates of the hazard function into estimates of the cumulative incidence functions.

```{r stanford-risk, eval = eval_cs4, echo = TRUE, dependson="stanford_fit", fig.cap="\\label{fig:cs4CIF} Plot of the cumulative incidence function for exposed and unexposed individuals."}
# Compute absolute risk curves
newdata <- data.frame(exposure = c(0, 1))
absrisk <- absoluteRisk(fit3,
  newdata = newdata,
  time = seq(0, 425, length.out = 100)
)

@@

@@

@@
